---
title: "Basic concepts and terminology"
description: |
  Chapter 2-1
# author:
#   - name: "Marcel Raab"
#     url: http://marcelraab.de
#     affiliation: University of Mannheim
#     affiliation_url: https://www.uni-mannheim.de
#   - name: "Emanuela Struffolino"
#     url: https://www.wzb.eu/en/persons/emanuela-struffolino
#     affiliation: WZB Berlin
#     affiliation_url: https://www.wzb.eu
# date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval=TRUE, echo=TRUE)

options("kableExtra.html.bsTable" = T)


# load data required for this subchapter
load("data/2-3_BasicDescription.RData")

# (down)load required packages using pacman
source("source/LoadInstallPackages.R")

```

<style>
.pre-scrolly {
  max-height: 250px;
  overflow-y: auto;
}
</style>

In chapter 2-1, we introduce different notations of sequence data using example data on family biographies from age 18 to 40. The data come from a sub-sample of the German Family Panel - pairfam. For further information on the study and on how to access the full scientific use file see [here](https://www.pairfam.de/en/).

## Defining a state sequence object

We generated the example dataset in Stata. Next to the sequence variables it comprises a few additional variables which will be used to analyze the sequences in later chapters.

We import the data to R using the `read_dta` function from the `haven` package and inspect the names of the imported variables.

```{r, eval = FALSE}
# import data
family <- read_dta("PartnerBirthbio.dta")
```

<div class='pre-scrolly'>
```{r, layout="l-body-outset"}
# view variable names
names(family)
```
</div>

The sequence variables begin with the prefix `state`. The data comprise 264 sequence variables per person (wide data format). These variables include monthly information on family biographies covering a period of 22 years.
Haven imports them as numeric variables with labels attached to them (`class = "haven_labelled"`). With the follwing commands we can take a look at the labels.

```{r}
str(family$state1) 
attributes(family$state1)$labels
```

The first examples in the book are based on sequences with a reduced alphabet only distinguishing partnership states. The following code generates a `data.frame` containing the recoded sequence variables using `dplyr`.

```{r, eval=FALSE}
# extracting and recoding the sequence variables (which all start with state)
# recode to reduced state space capturing partnership status only 
seqvars.partner <- family %>%
  select(starts_with("state")) %>%
  mutate_all(~(case_when(
    . < 3 ~ 1,            # Single
    . %in% c(3,4) ~ 2,    # LAT
    . %in% c(5,6) ~ 3,    # Cohabiting
    . > 6 ~ 4,)))         # Married
```

Then we define two vectors storing the long and short labels for the states in the newly defined alphabet. Once that is done, we can define the data as a state sequence object. Most TraMineR functions for analyzing sequences require the data to have this format.

```{r}

shortlab.partner <- c("S", "LAT", "COH", "MAR")
longlab.partner <-  c("Single", "LAT", "Cohabiting", "Married")

# create state sequence object
partner.month.seq <- seqdef(seqvars.partner,
                            labels = longlab.partner,
                            states = shortlab.partner,
                            weights = family$weight40)

```

Note that the `seqdef` function can include many more optional arguments. Some of these arguments - most importantly `cpal` - affect the appearance of state sequence plots rendered with `seqplot` or `seqplot.rf`. We cover the definition of color palettes on two separate pages ([definition of color palettes](rChapter2-4_color.html); [definition of grayscale palettes](rChapter2-4_grayscale.html))  

\  

## Sequence data notation 

In chapter 2-1, we introduce different notations for printing sequences. The following commands print the sequences in *STS*, *DSS*, and *SPS* format.


```{r}
print(partner.month.seq[8, ], format = "STS")
seqdss(partner.month.seq[8, ])
print(partner.month.seq[8, ], format = "SPS")
```

