---
title: "2.5 Advanced Description"
description: |
  Assessing sequence complexity and quality
# author:
#   - name: "Marcel Raab"
#     url: http://marcelraab.de
#     affiliation: University of Mannheim
#     affiliation_url: https://www.uni-mannheim.de
#   - name: "Emanuela Struffolino"
#     url: https://www.wzb.eu/en/persons/emanuela-struffolino
#     affiliation: WZB Berlin
#     affiliation_url: https://www.wzb.eu
# date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

options("kableExtra.html.bsTable" = T)


# load data required for this subchapter
load("data/2-3_BasicDescription.RData")

# (down)load required packages using pacman
source("0-0_LoadInstallPackages.R")

```

## Sequencing – counting transitions and subsequences

The chapter starts with two tiny sequences that were constructed to illustrate differences between counting the number of transitions and the number of subsequences. The sequence data are constructed with the following code:

```{r, eval=TRUE, echo=TRUE}
seqX <- c("S","LAT","COH","MAR")
seqY <- c("S","LAT","COH","S")

ex1.seq <- seqdef(rbind(seqX,seqY), alphabet = seqX)
```

The number of transitions can be obtained with the `seqtransn` function, the number of **distinct** subsequences is computed with `seqsubsn`. Both functions are part of the `TraMineR` package..

```{r, eval=TRUE, echo=TRUE}
# Number of transitions
seqtransn(ex1.seq)

# Number of subsequences
seqsubsn(ex1.seq)
```

In Table 2-9 we show all distinct subsequences extracted from *Sequence x* (`seqx`). The subsequences can be extracted by using the `combn` function. The functions extracts only subsequences of given length each time it is executed. In the following loop we specify `rev(seq_along(seqX))` to extract subsequences of length 4 to 1. The extracted subsequences are stored as data.frames in the resulting list `subseqs`. In the next step, we put all these subsequences into one data frame using `bind_rows(subseqs)`. Then we remove all duplicates using `distinct` and add an empty row for the empty subsequence $\lambda$. The resulting dataset can be nicely printed in the console with `print(seqdef(subseqs), format = "SPS")`. 


```{r, eval=TRUE, echo=TRUE}
# Extract & display all possible subsequences of Sequence x
subseqs <- vector(mode = "list", length = length(seqX))

for (i in rev(seq_along(seqX))) {
  subseqs[[i]] <- as.data.frame(t(combn(seqX, i)))
}

# Store all distinct subsequences in one dataset
subseqs <- bind_rows(subseqs)
subseqs <- distinct(subseqs)

#Add and fill empty row
subseqs <- add_row(subseqs, .before = 1)
subseqs[1,1] <- "-"

#Print subsequences
seqdss(seqdef(subseqs))
```


<!--
seqtransn(ex1.seq, norm = TRUE) # normalized
seqsubsn.max <- rep(alphabet(ex1.seq), length.out = max(seqlength(ex1.seq)))
seqsubsn.max <- seqsubsn(seqdef(t(seqsubsn.max)))
(log2(seqsubsn(ex1.seq))-1)/(log2(rep(seqsubsn.max,nrow(ex1.seq)))-1) # normalized
-->

To be continued...
